Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-27 ms/S2_appendix.tex
update fig S4 caption, etc

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{../figures/fig_S2_4.pdf}
\caption{{\bf Effects of contact heterogeneity and vital dynamics.}
This figure matches Fig 4 in main text, but displays the
summary statistics of the modified explicit models (models with contact heterogeneity and vital dynamics).
For each model, its summary statistics are plotted on the univariate plot using different shapes.
All models were simulated using the baseline parameters.
}
\label{fig:fig3aug}
\end{figure}

>>>>

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{../figures/fig_S2_4.pdf}
\caption{{\bf Effects of contact heterogeneity and vital dynamics.}
This figure matches Fig 4 in main text, but adds results
for model variants with contact heterogeneity and vital dynamics.
The violin plots, copied from Fig 4 in the main text, show the
distribution of outcomes across the Latin hypercube sample
for SIS model without contact heterogeneity. The points show
the outcome of various models using the baseline parameters
from Table 1: (\emph{circle}) unmodified 
models (SIS, homogeneous); (\emph{triangle}) models with vital dynamics; 
(\emph{square}) models with contact heterogeneity.
}
\label{fig:fig3aug}
\end{figure}

### 2017-01-27 ms/ms.tex
update fig S4 caption, etc

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate with many different model
structures. For the most part, our models are too complex to easily
derive analytical correspondences among them. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
(intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While $\rzero$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than $\rzero$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while $\rzero$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate in 
disease incidence. \todo{prevalence?}

@@

Finally, incorporating additional realism to the model, i.e.
combining heterogeneity with all four basic contact structures or
allowing for vital dynamics rather than assuming an SIS model,
leads to only small differences in the conclusions stated so far
(\figurename~??). Relative to our baseline SIS assumption, the
effect of adding vital dynamics is to delay the virulence peak
slightly and increase both the peak and equilibrium virulence.
The changes are small, however: across all models, the
maximum increase in time until
the virulence peak is 40 years (for the instswitch+epc model),in 
the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
\Lspvl\ is 0.4 (pairform). The changes in the most realistic
model (pairform+epc) are considerably smaller: 
an increase of 2 years vs. a decrease of 13 years in the
time to the virulence peak for the
models with vital dynamics and heterogeneity, respectively;
an increase of 0.1 units vs. a decrease of 0.01 units in peak \Lspvl;
and an increase of 0.145 units vs. a decrease of 0.03 units
in equilibrium~\Lspvl. Thus, while we can never rule out the
possibility of some higher-order interaction among epidemiological
phenomena leading to significant changes in our conclusions,
we are reasonably confident that the results reported here are
robust to additional complexities.

@@

%In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
%actually more true to real-world HIV epidemiological dynamics than a
%model with realistic partnership durations that neglects extra-pair
%contact \cite{herbeck2016evolution}. 
Taking further steps to make the model even more realistic
might add further structure,
making the random-mixing model predictions less accurate. For
example, our model forms partnerships randomly, and assumes that
extra-pair contact is randomly mixing across the population;
one could instead model extra-pair contact as arising from
multiple concurrent partnerships (some, such as contact with sex
workers, of very short duration) and/or more structured partnership
formation (by age, ethnicity, or behaviour group). The effects of
other realistic complications such as explicit modeling of two
sexes (both in contact structure and differential transmission
probabilities), temporal and spatial variation in epidemic processes,
or presence of genetic variation in hosts are harder to predict.

>>>>

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate with many different model
structures. For the most part, our models are too complex to easily
derive analytical correspondences among them. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
(intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While $\rzero$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than $\rzero$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while $\rzero$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate in 
disease prevalence.

@@

Finally, incorporating additional realism to the model, i.e.
combining heterogeneity with all four basic contact structures or
allowing for vital dynamics rather than assuming an SIS model,
leads to only small differences in the conclusions stated so far
(\figurename~S4). Relative to our baseline SIS assumption, the
effect of adding vital dynamics is to delay the virulence peak
slightly and increase both the peak and equilibrium virulence.
The changes are small, however: across all models, the
maximum increase in time until
the virulence peak is 40 years (for the instswitch+epc model),in 
the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
\Lspvl\ is 0.4 (pairform). The changes in the most realistic
model (pairform+epc) are considerably smaller: 
an increase of 2 years vs. a decrease of 13 years in the
time to the virulence peak for the
models with vital dynamics and heterogeneity, respectively;
an increase of 0.1 units vs. a decrease of 0.01 units in peak \Lspvl;
and an increase of 0.145 units vs. a decrease of 0.03 units
in equilibrium~\Lspvl. Thus, while we can never rule out the
possibility of some higher-order interaction among epidemiological
phenomena leading to significant changes in our conclusions,
we are reasonably confident that the results reported here are
robust to additional complexities.

@@

Taking further steps to make the model even more realistic
might add further structure,
making the random-mixing model predictions less accurate. For
example, our model forms partnerships randomly, and assumes that
extra-pair contact is randomly mixing across the population;
one could instead model extra-pair contact as arising from
multiple concurrent partnerships (some, such as contact with sex
workers, of very short duration) and/or more structured partnership
formation (by age, ethnicity, or behaviour group). The effects of
other realistic complications such as explicit modeling of two
sexes (both in contact structure and differential transmission
probabilities), temporal and spatial variation in epidemic processes,
or presence of genetic variation in hosts are harder to predict.
